% !TEX root = ../main.tex
%
\chapter{Conclusion}
\label{sec:conclusion}
\section{Summary}
\label{sec:conclusion:summary}
As LLMs are increasingly deployed in high-stakes domains, their tendency to 
hallucinate plausible but factual incorrect information remains a critical
vulnerability. Standard uncertainty quantification methods rely on single,
precise probability distributions. By mathematically conflating benign
linguistic diversity (aleatoric uncertainty) and with a genuine lack of
knowledge (epistemic uncertainty), models frequently appear misleadingly 
confident in their hallucinations.

To address this theoretical limitation, this thesis introduced
\textbf{Credal Semantic Entropy}, a novel framework grounded in the theory 
of imprecise probabilities. We hypothesized that true epistemic uncertainty
cannot be reliably captured by a single fixed decoding strategy. Instead,
we shifted the locus of uncertainty estimation from a single 
"point-estimate" to a \textbf{credal set} - a convex set of plausible 
probability distributions. Our methodology constructed this set by 
systematically varying the generation temperature, utilizing the statistical
concept of \textbf{relative likelihood} to filter out implausible 
temperatures via an $\alpha$-cut. We then clustered the generated sequences
by semantic meaning using an auxiliary Natural Language Inference
model, deriving robust upper and lower probability bounds for every
fundamental meaning.

We empirically evaluated our approach on the conversational Question 
Answering benchmark, CoQA, using the OPT-6.7B model. Across multiple 
temperature grids (20-step and 50-step) and varying $\alpha$ thresholds,
we compared the upper semantic entropy of our credal set against established
single-point baselines.

Our empirical results revealed that while the credal framework is 
theoretically sound, it faces significant practical challenges. Under our 
tested configurations, strong standard baselines outperformed the proposed
credal semantic entropy in detecting hallucinations (measured via AUROC).
%We observed that the performance of the credal approach is highly sensitive to the $\alpha$ threshold, overly restrictive cuts collapse the probability intervals, neutralizing the benefits of the set-based approach.

Ultimately, this thesis successfully establishes a formal mathematical 
bridge between LLM decoding hyperparameters and imprecise probabilities. 
It demonstrates that future research must continue to explore interval-based
uncertainty, particularly by refining semantic clustering techniques,
isolating the purely epistemic component of the credal set, and optimizing
computational efficiency.



\section{Limitations}
\label{sec:conclusion:limitations}
While the proposed credal semantic entropy framework provides a rigorous,
theoretically grounded approach to hallucination detection, this study is
subject to several practical and methodological limitations.
The most significant limitation of our approach is the high computational
cost incurred during inference.
Because of these computational demands and the time constraint of this 
thesis, our empirical evaluation was restricted in scope. We evaluated the
method on a subset of approximately 2400 queries from a single dataset 
(CoQA \parencite{guo2021abgcoqa}). Evaluating the method across a wider 
variety of domains and with
more granular temperature grids was computationally prohibitive.
Furthermore all experiments were conducted on the OPT-6.7B
\parencite{zhang2022opt} model. This
specific architecture was selected because it is an open-weights 
("white-box") model, granting full access to token-level logits required 
for our calculations, while also being small enough to run on accessible 
hardware without incurring the prohibitive API costs associated with
proprietary models. However, it remains an open question whether the
uncertainty dynamics observed in a 6.7-billion parameter model perfectly
mirrors those of massive, instruction-tuned foundation models (e.g. GPT-4
\parencite{openai2024gpt4technicalreport}).


\section{Future Work}
\label{sec:conclusion:future}
While this thesis establishes a rigorous baseline for credal semantic 
entropy using temperature scaling, several avenues remain for refining 
the methodology and extending its application. We outline three primary
directions for future research.

\subsection{Generalization Across Domains and Architecture}
\label{sec:conclusion:future:generalization}
Our experiments focused on the CoQA dataset using the OPT-6.7B model.
To validate the universality of the credal set approach, future work
should assess performance on:

\begin{itemize}
    \item \textbf{diverse benchmarks} such as TriviaQA 
	\parencite{joshi2017triviaqalargescaledistantly} and 
	SQuAD \parencite{rajpurkar2018knowdontknowunanswerable};
    \item \textbf{safety-critical domains} like healthcare 
	(MedQA \citealp{jin2020diseasedoespatienthave}) or legal 
	analysis, where the cost of epistemic failure is significantly 
	higher;
    \item \textbf{state-of-the-art foundation models} such as Gemini 
	\parencite{geminiteam2025geminifamilyhighlycapable} or 
	GPT4.
\end{itemize}

\subsection{Multi-Dimensional Decoding Ensembles}
\label{sec:conclusion:future:decoding}
In this work, we constructed the credal set solely by varying the 
\textit{temperature} parameter. However, the decoding space of modern 
LLMs is high-dimensional. A natural extension is to define the credal 
set over a broader range of decoding hyperparameters, including:
\textbf{Nucleus (Top-p) Sampling} (Varying the cumulative 
probability threshold \parencite{holtzman2020curiouscaseneuraltext}),
\textbf{Top-k Sampling} (Varying the vocabulary truncation size 
\parencite{fan2018hierarchicalneuralstorygeneration}), and
\textbf{Beam Search} (comparing stochastic sampling against deterministic 
beam search paths \parencite{freitag-al-onaizan-2017-beam}).
By defining the $\alpha$-cut over this joint hyperparameter space, we 
could capture a more holistic view of the model's inherent uncertainty.

\subsection{Methodological Extensions: Ensembles and Black-Box Settings}
\label{sec:conclusion:future:black_box}
Finally, the relative likelihood framework could be extended beyond 
single-model decoding.
\textbf{Model Ensembles:} Instead of perturbing the temperature of a 
single model, one could construct a credal set using an ensemble of 
different model checkpoints or architectures, effectively bridging 
the gap between our approach and the original parameter-space credal 
prediction \parencite{löhr2025credalpredictionbasedrelative}.
\textbf{Discrete Semantic Entropy:} Currently, our method requires 
access to token-level log-probabilities (White-Box). To extend this 
method to closed-source "Black-Box" APIs (e.g., Claude 
\parencite{claude3}, GPT-4 \parencite{openai2024gpt4technicalreport}), 
future work could integrate \textbf{Discrete Semantic Entropy} 
\parencite{kuhn2024semanticentropy}, approximating the cluster 
probabilities using purely text-based sampling without access to 
internal logits.
