% !TEX root = ../main.tex
%
\chapter{Introduction}
\label{sec:intro}

\section{Motivation}
\label{sec:intro:motivation}
Large Language Models (LLMs) have fundamentally transformed the  
landscape of Artificial Intelligence, demonstrating unprecedented
capabilities in natural language understanding, reasoning and 
generation 
\parencite{brown2020languagemodelsfewshotlearners, wei2023chainofthoughtpromptingelicitsreasoning}.
% PROOF:
LLMs are increasingly deployed in critical domains ranging 
from healthcare \parencite{10.1001/jama.2024.21700} 
to legal analysis \parencite{10850911}. 
However, despite their fluency, 
these models suffer from a critical reliability failure:
the tendency to generate plausible-sounding but factual incorrect 
statements, a phenomenon widely known as hallucinations 
\parencite{Huang2023ASO}.

For high-stakes applications, accuracy alone is insufficient. A 
system must also be self-aware of its limitations 
\parencite{bommasani2022opportunitiesrisksfoundationmodels}. 
Unfortunately, modern LLMs 
are frequently "confidently wrong", assigning high probability scores 
to their hallucinations due to poor calibration 
\parencite{guo2017calibrationmodernneuralnetworks, 
desai2020calibrationpretrainedtransformers} 
and the nature of their training objectives \parencite{Ji_2023}.

Standard uncertainty metrics, such as predictive entropy
\parencite{malinin2021uncertaintyestimationautoregressivestructured}
or lexical similarity 
\parencite{fomicheva2020unsupervisedqualityestimationneural}, 
often fail to detect these errors because they 
rely on a single, precise probability distribution. This 
"point-estimate" approach mathematically conflates two distinct 
sources of uncertainty \parencite{H_llermeier_2021}: 
\textit{aleatoric uncertainty} 
(the natural diversity of valid human language) 
and \textit{epistemic uncertainty} (the model's lack of knowledge). 
When an LLM distributes its 
probability mass across five different ways of saying the correct
answer, standard metrics flag this as high uncertainty, confusing 
benign linguistic flexibility with factual ignorance.

\section{Problem Statement}
\label{sec:intro:problem_statement}
The core problem addressed in this thesis is the unreliability of
single-point probability estimates for detecting hallucinations in
free-form Question Answering.

Current state-of-the-art methods, such as Semantic entropy
\parencite{kuhn2023semanticuncertaintylinguisticinvariances},
attempt to mitigate this by grouping lexically diverse answers
into semantic clusters. However, even Semantic Entropy relies on a
single, static probability distribution generated at a fixed 
temperature. If this underlying distribution is miscalculated,
i.e., peaked around an incorrect fact, the metric will fail to
signal a warning.
To robustly detect hallucinations, we need a framework that moves 
beyond precise probabilities and considers a set of plausible 
model states.

\section{Contributions}
\label{sec:intro:contributions}
This thesis proposes a novel methodology for quantifying epistemic
uncertainty by bridging the gap between imprecise probabilities
and LLM decoding.

Drawing on the theory of \textit{Credal Prediction based on Relative 
Likelihood} \parencite{lĂ¶hr2025credalpredictionbasedrelative}, we 
introduce a statistically rigorous method to construct a credal set 
of semantic distributions. Our specific contributions are as follows: 

\begin{itemize}
    \item \textbf{Valid Temperature Selection via Relative Likelihood}:
	We propose a method to identify a statistically valid range of 
	decoding temperatures. By treating temperatures as a 
	hyperparameter and applying an $\alpha$-cut, we systematically 
	discard implausible distributions.

    \item \textbf{Construction of Semantic Credal Sets}:
	We extend the concept of Semantic Entropy to operate over 
	Interval-Valued Probabilities. By computing the minimum and 
	maximum probability of semantic clusters across valid 
	temperatures, we construct a convex credal set that captures
	the model's true epistemic ambiguity. 

    \item \textbf{Empirical Validation}:

\end{itemize}


\section{Thesis Structure}
\label{sec:intro:structure}
The remainder of this thesis is organized as follows: 
\begin{itemize}
	\item \textbf{Chapter 2 Background}:
	    establishes the theoretical foundations 
	    necessary for this work. It reviews the architecture of 
	    LLMs, explains autoregressive decoding 
	    strategies, and introduces the mathematical framework of 
	    imprecise probabilities and credal sets.

	\item \textbf{Chapter 3 Related Work}:
	    surveys the current landscape of uncertainty 
	    quantification. It contrasts traditional point-estimate 
	    methods with recent advancements in semantic consistency, 
	    specifically focusing on Semantic Entropy as the primary 
	    baseline for our research.

	\item \textbf{Chapter 4 methodology}:
	    formalizes our novel contribution: 
	    \textit{Credal Semantic Entropy}. It details the algorithm 
	    for selecting valid temperatures via relative likelihood 
	    and deriving robust uncertainty bounds from 
	    interval-valued probability distributions.

	\item \textbf{Chapter 5 Experiments}:
	    presents the experimental setup and results, 
	    analyzing the performance of our method against strong 
	    baselines.

	\item \textbf{Chapter 6 Conclusion}:
\end{itemize}
