% !TEX root = ../main.tex
%
\chapter{Background}
\label{sec:background}
This chapter establishes the theoretical and methodological 
foundations necessary to understand the contributions of this thesis. 
Our work lies at the intersection of Natural Language Processing (NLP)
and Uncertainty Quantification (UQ). 
We first establish the mathematical principles of uncertainty in
standard machine learning, followed by the probabilistic 
mechanisms of modern language models and finally how these 
domains unite in our methodology.


\section{Uncertainty in Machine Learning}
\label{sec:background:uncertainty}
As LLMs are increasingly deployed in real-world and high-stakes
applications, for example in healthcare
\parencite{10.1001/jama.2024.21700},
their reliability becomes a critical concern.
A fundamental challenge with deep neural networks, including LLMs,
is that they are prone to overconfidence. They frequently assign
high probabilities to incorrect predictions
\parencite{guo2017calibrationmodernneuralnetworks}.
Consequently, evaluating a model's performance requires not only
measuring its accuracy but also quantifying its \textbf{uncertainty}.


\subsection{Aleatoric and Epistemic Uncertainty}
\label{sec:background:uncertainty:aleatoric_epistemic}
In the field of machine learning, uncertainty can be 
categorized into two distinct sources: aleatoric and epistemic
\parencite{kendall2017uncertaintiesneedbayesiandeep, H_llermeier_2021}.

\textbf{Aleatoric Uncertainty} arises from the inherent complexity,
noise, or randomness in the underlying data generation process. It
is irreducible, meaning providing the model with more training data 
will not eliminate this uncertainty
\parencite{kendall2017uncertaintiesneedbayesiandeep, H_llermeier_2021}.
In a standard classification task, this might occur when two 
distinct classes share highly similar features, making perfect
separation impossible regardless of the model's capacity.


\textbf{Epistemic uncertainty}, on the other hand, stems from a lack
of knowledge or information about the best model.
This uncertainty occurs when a model encounters out-of-distribution
(OOD) inputs or queries about facts absent from its training data.
Unlike aleatoric uncertainty, epistemic uncertainty is reducible.
Theoretically it can be minimized by providing the model with more 
comprehensive training data
\parencite{kendall2017uncertaintiesneedbayesiandeep, H_llermeier_2021}.


\subsection{Standard Metrics: Shannon Entropy and Maximum Softmax probability (MSP)}
\label{sec:background:uncertainty:standard_metrics}
To quantify these theoretical concepts, we rely on established
mathematical metrics. For a standard classification model 
predicting a class $Y$ given input $\mathbf{x}$, two fundamental 
measures of predictive confidence are \textbf{Shannon Entropy}
\parencite{https://doi.org/10.1002/j.1538-7305.1948.tb01338.x} and 
\textbf{Maximum Softmax probability (MSP)} 
\parencite{hendrycks2018baselinedetectingmisclassifiedoutofdistribution}.

\textbf{Shannon Entropy} 
quantifies the fundamental uncertainty inherent in the entire
probability distribution over all possible classes 
$y \in \mathcal{Y}$.
For a discrete random variable $Y$ with a probability 
distribution
$P(Y \mid \mathbf{x})$, the Shannon Entropy $H$ is defined as
\parencite{https://doi.org/10.1002/j.1538-7305.1948.tb01338.x}:
\begin{equation}
    H(Y) = - \sum_{y \in \mathcal{Y}} P(y \mid \mathbf{x}) \log P(y \mid \mathbf{x})
    \label{eq:shannon_entropy}
\end{equation}

Entropy provides a scalar summary of the distribution's shape.
\begin{itemize}
    \item \textbf{Low Entropy} ($H \rightarrow 0$): The distribution
	is highly concentrated (peaked) on one or a few tokens, 
	indicating that the model is highly certain about its 
	next prediction.

    \item \textbf{High Entropy} ($H \rightarrow \log |\mathcal{V}|$):
	The distribution is flat and uniform across the vocabulary,
	indicating maximum uncertainty.
\end{itemize}


\textbf{Maximum Softmax Probability (MSP)} \\
While Shannon Entropy evaluates the uncertainty across the entire
label space $\mathcal{Y}$, the most ubiquitous measure of
point-prediction confidence in deep learning is the Maximum Softmax
Probability. For a given input $\mathbf{x}$, MSP simply takes the
probability of the most likely predicted class $\hat{y}$ as the 
model's absolute confidence score:

\begin{equation}
    MSP(\mathbf{x}) = \max_{y \in \mathcal{Y}} P(y \mid \mathbf{x})
    \label{eq:msp}
\end{equation}

MSP provides a straightforward scalar value ranging from 
$\frac{1}{|\mathcal{Y}|}$ (complete uncertainty) to 1.0 
(absolute certainty).

\textbf{Limitations of Standard Metrics} \\
Shannon Entropy and MSP are highly effective metrics, provided 
that the underlying probability distribution $P$ is perfectly 
calibrated and accurately reflects the true state of the world. 
However, these metrics inherently assume that the model's output
distribution is a single, precise, and entirely correct representation
of reality 
\parencite{guo2017calibrationmodernneuralnetworks, Gneiting01032007}. 

As demonstrated by \textcite{guo2017calibrationmodernneuralnetworks},
this assumption of a ``precise probability'' often breaks down in modern
deep learning models, which frequently suffer from 
severe overconfidence and miscalibration. Furthermore, standard precise
probabilities mathematically struggle to distinguish between
inherent data noise (aleatoric) and a fundamental lack of 
knowledge (epistemic) \parencite{H_llermeier_2021}. 
This failure to capture true epistemic misconfidence
necessitates more robust mathematical frameworks, such as
imprecise probabilities and credal sets\parencite{Walley1991-WALSRW}.


\section{Imprecise Probabilities and Credal Sets}
\label{sec:background:credal_sets}
Traditional probability theory operates on the assumption that all
forms of uncertainty can be adequately represented by a single, 
precise probability distribution \parencite{Walley1991-WALSRW}.
Under this paradigm, an agent or model is required to assign an exact
numerical probability to every possible event, regardless of the 
amount of evidence available. While computationally convenient,
this precise framework struggles to model severe epistemic
uncertainty. When a model lacks sufficient training data or 
encounters an ambiguous query, forcing it to produce a single 
probability distribution often results in arbitrary or misleadingly
confident predictions. Precise probabilities mathematically conflate
the lack of evidence (epistemic uncertainty) with known stochastic
noise (aleatoric uncertainty), masking the model's true state of 
ignorance \parencite{H_llermeier_2021}. Specifically, a single, flat
probability distribution cannot distinguish between an inherently
ambiguous input and a completely novel one, as both scenarios yield
the same low-confidence prediction.

\subsection{Introduction to Credal Sets}
\label{sec:background:credal_sets:intro}
To address the limitations of precise probabilities, the framework of
\textbf{imprecise probabilities} replaces the single distribution
with a set of plausible distributions
\parencite{Levi1980-LEVTEO-7, Walley1991-WALSRW}.
A closed, convex set of probability distributions is known as a 
\textbf{credal set} \parencite{COZMAN2000199}, denoted by 
$\mathcal{K}$. 


\subsection{The Geometry of Credal Sets}
\label{sec:background:credal_sets:geometry}
A credal set $\mathcal{K}$ is geometrically represented as a closed, 
convex subset of the standard probability simplex $\Delta$. 
For a classification task with $K$ possible classes 
$c_1, c_2, \dots, c_K$ (or semantic clusters), 
the standard simplex $\Delta_K$ is defined as
\parencite{https://doi.org/10.1002/bimj.4710300705}:

\begin{equation}
    \Delta_K = \left\{ P \in \mathbb{R}^K \mid 
	\forall k: P_k \geq 0, \sum_{k=1}^K P_k = 1
	\right\} 
    \label{eq:simplex}
\end{equation}

A probability distribution is a single point within this simplex. 
In contrast, a credal set represents a region of points.

\textbf{Interval-Valued Probabilities} \\
One of the most common and practical ways to define this region is  
through \textbf{probability intervals}. Instead of assigning a precise
value $P(c_k \mid \mathbf{x})$, we assign a lower bound $\underline{P}(c_k \mid \mathbf{x})$ and an
upper bound $\overline{P}(c_k \mid \mathbf{x})$ to each class $c_k$
\parencite{article}.

The credal set induced by these intervals is the set of all 
distributions consistent with these bounds:

\begin{equation}
    \mathcal{K} = \left\{ P \in \Delta_K \mid \underline{P} (c_k \mid \mathbf{x}) 
	\leq P(c_k \mid \mathbf{x}) \leq \overline{P}(c_k \mid \mathbf{x}), \sum_{k=1}^{K} P(c_k \mid \mathbf{x}) = 1 
	\right\}
    \label{eq:probability_interval_credal_set}
\end{equation}

Geometrically, these linear constraints form a \textbf{convex polytope}
within the simplex
\parencite{article}.

If the intervals are wide (high epistemic uncertainty), the polytope
covers a large volume of the simplex. If the intervals are narrow
(low uncertainty), the polytopes shrinks toward a single point.
This geometric interpretation is fundamental to understanding how
we quantify uncertainty in Chapter \ref{sec:methodology}.


\subsection{Computing Uncertainty on Credal Sets}
\label{sec:background:credal_sets:uncertainty}
To quantify the uncertainty of a model's prediction 
using a credal set, we must extend standard metrics like 
Shannon Entropy to operate over sets of distributions.
Since we have a set of probability distributions $\mathcal{K}$, 
we can determine the \textbf{lower entropy} and  
\textbf{upper entropy} \parencite{ABELLAN2005235} 
by finding the minimum and maximum entropy 
values across the set \parencite{Abellán01022006, H_llermeier_2021}:

\begin{equation}
    H_*(\mathcal{K}) = \min_{P \in \mathcal{K}} H(P), 
	\quad H^*(\mathcal{K}) = \max_{P \in \mathcal{K}} H(P)
    \label{eq:lower_upper_entropy}
\end{equation}

We can use these bounds to quantify total uncertainty $U(\mathcal{K})$,
aleatoric uncertainty $AU(\mathcal{K})$ and epistemic uncertainty 
$EU(\mathcal{K})$ \parencite{Abellán01022006, H_llermeier_2021}:

\begin{equation}
    U(\mathcal{K}) = AU(\mathcal{K}) + EU(\mathcal{K})
    \label{eq:credal_total_uncertainty}
\end{equation}

\begin{equation}
    H^*(\mathcal{K}) = H_*(\mathcal{K}) + 
	\left(H^*(\mathcal{K}) - H_*(\mathcal{K})\right)
    \label{eq:credal_uncertainty}
\end{equation}

Here, upper entropy represents total uncertainty \parencite{ABELLAN2005235},
lower entropy represents aleatoric uncertainty and the difference 
between them is used as a measure for epistemic uncertainty. 

\section{Large Language Models and Generation}
\label{sec:background:llms}
Having established the framework for quantifying uncertainty,
we now turn to the models under investigation.
Large Language Models (LLMs) represent a class of 
deep learning models designed to process, 
understand, and generate human language. 
At their core, modern LLMs are built upon the 
Transformer architecture,
utilizing mechanisms such as self-attention to model 
complex dependencies between words across long sequences 
\parencite{NIPS2017_3f5ee243, naveed2024comprehensiveoverviewlargelanguage}.
They demonstrate capabilities in reasoning 
\parencite{wei2023chainofthoughtpromptingelicitsreasoning} 
and coding \parencite{tian2023chatgptultimateprogrammingassistant}.
Their fundamental operation remains probabilistic. They 
predict the likelihood of the next token in a sequence
based on the preceding context
\parencite{NIPS2017_3f5ee243, bahdanau2016neuralmachinetranslationjointly}.


\subsection{Transformers and Next-token prediction}
\label{sec:background:llms:transformers}
The primary objective of an autoregressive LLM is to model
the probability distribution of a target sequence of tokens
$\mathbf{y} = (y_1, y_2, \dots, y_T)$ given an input context or
prompt $\mathbf{x}$. The model decomposes the joint probability of
the sequence into a product of conditional probabilities
using the chain rule of probability 
\parencite{bahdanau2016neuralmachinetranslationjointly}:
\begin{equation}
    P(\mathbf{y} | \mathbf{x}) = \prod_{t=1}^{T} P(y_t \mid \mathbf{y}_{<t}, \mathbf{x})
    \label{eq:sequence_probability}
\end{equation}
Where $y_t$ is the token at step $t$ and $\mathbf{y}_{<t}$ represents
all preceding tokens $y_1, \dots, y_{t-1}$ in the generated sequence.
To compute these probabilities, the Transformer processes the 
input $\mathbf{x}$ and the generated history $\mathbf{y}_{<t}$ to produce a 
vector of non-normalized scores, known as logits, for every 
token in the vocabulary $\mathcal{V}$. Let $z_i$ be the logit 
corresponding to the $i$-th token $v_i \in \mathcal{V}$. The 
model converts these logits into a valid probability
distribution using the Softmax function
\parencite{10.1007/978-3-642-76153-9_28, NIPS2017_3f5ee243}:
\begin{equation}
    P(y_t = v_i \mid \mathbf{y}_{<t}, \mathbf{x})
	= \frac{\exp(z_i)}{\sum_{j=1}^{|\mathcal{V}|} \exp(z_j)} 
	\quad \quad \forall i \in {1, \dots, |\mathcal{V}|}
    \label{eq:logit_softmax}
\end{equation}
This process is repeated iteratively, where the token 
selected at step $t$ is appended to the input for step $t + 1$
\parencite{NIPS2017_3f5ee243}.


\subsection{Decoding Strategies and the Role of Temperature}
\label{sec:background:llms:decoding}
Once the probability distribution over the Vocabulary is computed for 
the next token, a \textbf{decoding strategy} determines what token will
be selected. Common maximization-based strategies include 
\textbf{Greedy Decoding} or \textbf{Beam Search}.
\textbf{Greedy Decoding} is the simplest deterministic method. It 
selects the token with the highest probability mass at the current
time step $t$ \parencite{sutskever2014sequencesequencelearningneural}:
\begin{equation}
    y_t = \argmax_{y \in \mathcal{V}} P(y \mid \mathbf{y}_{<t}, \mathbf{x})
    \label{eq:greedy_decoding}
\end{equation}
While computationally efficient, this strategy often yields 
sub-optimal results. By focusing solely on local probability 
maximization, the strategy risks settling into local optima, 
effectively missing high-probability sequences hidden behind 
lower-probability initial tokens.
Furthermore, greedy decoding is known to produce repetitive and 
generic text, as it fails to account for the diversity and 
global coherence of the sequence
\parencite{wiseman-rush-2016-sequence, holtzman2020curiouscaseneuraltext}.

\textbf{Beam Search} extends Greedy Decoding by maintaining multiple 
hypotheses at each time step, rather than a single sequence. It 
introduces a hyperparameter $k$, known as the \textbf{beam width}, 
which determines the number of candidates to keep.
At each step $t$, the algorithm expands all $k$ candidates from the 
previous step by appending every possible next token from the 
vocabulary $\mathcal{V}$
\parencite{freitag-al-onaizan-2017-beam}.

Formally, let $\mathcal{B}_{t-1} = \{\mathbf{y}^{(1)}_{<t}, \dots, 
\mathbf{y}^{(k)}_{<t}\}$ be the set of $k$ hypothesis sequences 
generated up to step $t-1$. The algorithm proceeds in three steps
\parencite{freitag-al-onaizan-2017-beam}:
\begin{enumerate}
	\item \textbf{Expansion}: Each hypothesis in 
	    $\mathcal{B}_{t-1}$ is extended with every possible token
	    $v$ from the vocabulary $\mathcal{V}$, creating a candidate
	    set $\mathcal{C}_t$.

	\item \textbf{Scoring}: Each candidate sequence is assigned a 
	    score based on its cumulative log-probability. The score 
	    for a sequence $\mathbf{y}_{1:t}$ 
	    (representing $y_1,\dots,y_t$)
	    is defined recursively:
	    \begin{equation}
		\text{Score}(\mathbf{y}_{1:t})
		    = \text{Score}(\mathbf{y}_{1:t-1}) 
		    + \log P(y_t \mid \mathbf{y}_{<t}, \mathbf{x})
	    	\label{eq:Beam_Score}
	    \end{equation}

	\item \textbf{Selection (Pruning)}: The algorithm selects the
	    $k$ candidates with the highest scores to form the new
	    hypothesis set $\mathcal{B}_t$:
	    \begin{equation}
		\mathcal{B}_t = 
	    \operatorname*{arg-top-\textit{k}}_{\mathbf{y}_{1:t} \in \mathcal{C}_t}
	    \left(\text{Score}(\mathbf{y}_{1:t})\right)
	    	\label{eq:Beam_Selection}
	    \end{equation}
\end{enumerate}
This approach allows the model to recover from locally sub-optimal 
decisions if they lead to a highly probable global sequence. However,
Beam Search is computationally more expensive than greedy methods and, 
as noted by \textcite{holtzman2020curiouscaseneuraltext}, can still
suffer from repetitive loops or generic responses in open-ended
generation tasks.

\textbf{Sampling} represents a fundamental shift from 
maximization-based decoding strategies towards stochastic generation.
Instead of strictly selecting the token with the highest probability, 
the model randomly draws the next token $y_t$ from the conditional 
probability distribution $P(y_t \mid \mathbf{y}_{<t}, \mathbf{x})$.

Formally, at each time step $t$, the next token is sampled according 
to: 
\begin{equation}
    y_t \sim P(y \mid \mathbf{y}_{<t}, \mathbf{x})
    \label{eq:sampling}
\end{equation}
This stochasticity allows the model to generate more diverse and 
creative outputs for the same input. However, pure random sampling
from the full distribution can occasionally lead to incoherence.
If the ``tail'' of the distribution, that contains very low-probability
tokens, is sampled, it may disrupt the semantic flow or factual 
accuracy of the sequence.
To mitigate this risk while preserving diversity, two truncation 
techniques are commonly employed:
\begin{enumerate}
	\item \textbf{Top-k Sampling} 
	    \parencite{fan2018hierarchicalneuralstorygeneration}:
		The model samples from the top $k$ most probable
		tokens. This effectively truncates the unreliable
		tail of the distribution. 
	\item \textbf{Nucleus (Top-p) Sampling}
	    \parencite{holtzman2020curiouscaseneuraltext}:
		The model samples from the smallest set of tokens
		whose cumulative probability exceeds a threshold
		$p$ (e.g., $p = 0.9$). Unlike Top-k, which uses
		a fixed number of tokens, Top-p adapts dynamically.
		The set of candidates grows when the model is 
		uncertain (flat distribution) and shrinks when the 
		model is confident (peaked distribution).
\end{enumerate}
While these truncation methods improve coherence, they operate 
by removing options. A more continuous method for controlling the shape 
of the probability distribution, and the core mechanism for this 
thesis, is \textbf{Temperature Scaling}.

\textbf{Temperature Scaling} is a technique used to calibrate the
confidence of the model's predictions. Unlike Top-$k$ or 
Nucleus Sampling, which truncate the vocabulary, temperature
scaling modifies the probability distribution itself by rescaling 
the logits before the Softmax function is applied
\parencite{hinton2015distilling}.

Let $z_i$ be the logit for token $v_i$. We introduce a hyperparameter
$\tau$ > 0, called \textbf{temperature}. 
The probability of selecting token $v_i$ is then given by:
\begin{equation}
    P^\tau(y_t = v_i \mid \mathbf{y}_{<t}, \mathbf{x})
    = \frac{\exp(z_i/\tau)}{\sum_{j = 1}^{|\mathcal{V}|} \exp(z_j/\tau)}
    \label{eq:temperature_scaling}
\end{equation}
The value of $\tau$ controls the entropy of the output distribution.
When choosing a \textbf{low temperature}, $\tau < 1.0$, the 
distribution becomes sharper, i.e., more peaked. The difference 
between large and small logits is exaggerated, causing the model 
to become more confident and deterministic. In the limit, this 
approximates greedy decoding. 
Using a \textbf{high temperature}, $\tau > 1.0$, flattens the 
distribution towards a uniform distribution. This increases the 
likelihood of selecting lower-probability tokens, resulting in 
more diverse but also potentially more incoherent or hallucinatory
outputs. For the temperature $\tau = 1.0$, the distribution remains
unchanged, reflecting the model's original logits 
\parencite{hinton2015distilling}.

In standard generation tasks, $\tau$ is typically a fixed 
hyperparameter
tuned to balance diversity and equality. However, in this thesis, we 
interpret temperature variation as a mechanism to explore the model's 
epistemic uncertainty. 
% NOTE: Do I want to keep that sentence?


\section{Quantifying Uncertainty in LLMs}
\label{sec:background:quantifying_uncertainty_in_llms}
Applying the principles of uncertainty quantification to LLMs
presents unique challenges that do not exist in standard 
classification tasks. In classification, the label space is fixed
and mutually exclusive. In natural language generation, however,
the model must navigate a massive, interconnected vocabulary over
multiple autoregressive steps, making it difficult to disentangle
aleatoric from epistemic uncertainty.

In NLP, aleatoric uncertainty
is highly prevalent due to the inherent ambiguity of language.
Often, there are multiple semantically equivalent ways to express the
same underlying thought. For example, ``It is raining heavily'' and
``It is pouring outside'' describe the same state, distributing the 
model’s probability mass across different valid token sequences 
\parencite{kuhn2023semanticuncertaintylinguisticinvariances}.

Epistemic uncertainty arises when the model suffers from a genuine
lack of knowledge, such as when it faces out-of-distribution (OOD)
prompts or queries about facts absent from its training data
\parencite{kendall2017uncertaintiesneedbayesiandeep, H_llermeier_2021}.

\subsection{Negative Log-Likelihood in LLMs}
\label{sec:background:quantifying_uncertainty_in_llms:nll_in_llms}
While standard classification evaluates the likelihood of a 
single class prediction, evaluating an LLM requires quantifying
the model's confidence over an entire generated sequence.
For a given input context $\mathbf{x}$ and a target sequence
$\mathbf{y}$, the sequence likelihood is the product of the 
individual token probabilities, as defined by the chain rule
(Equation \ref{eq:sequence_probability}).

Because multiplying many small probabilities for long sequences
leads to numerical underflow, it is standard practice to compute 
the \textbf{Negative Log-Likelihood (NLL)} 
\parencite{Goodfellow-et-al-2016}.
Taking the negative logarithm transforms the product of the 
probabilities into a sum. Furthermore, to fairly compare sequences 
of different lengths, the NLL can be normalized by the total
number of tokens $T$, yielding the average NLL per token
\parencite{murray2018correctinglengthbiasneural}:

\begin{equation}
    \text{NLL}(\mathbf{y} \mid \mathbf{x}) 
    = -\sum_{t=1}^{T} \log P(y_t \mid \mathbf{y}_{<t}, \mathbf{x})
    \label{eq:nll}
\end{equation}

\begin{equation}
    \text{average NLL}(\mathbf{y} \mid \mathbf{x}) 
    = -\frac{1}{T} \sum_{t=1}^{T} \log P (y_t \mid \mathbf{y}_{<t}, \mathbf{x})
    \label{eq:average_nll}
\end{equation}
A lower NLL indicates that the model assigns a higher
probability to the target sequence, implying higher confidence.
\section{Relative Likelihood and Valid Temperature Sets}
\label{sec:background:rel_ll_and_temp_sets}
To systematically construct a credal set, we need a rigorous criterion
to distinguish between plausible and implausible probability
distributions. Following recent advances in credal prediction
\parencite{löhr2025credalpredictionbasedrelative}, 
we achieve this by employing the statistical concept of 
\textbf{relative likelihood} 
\parencite{largesampledistribution, cox1979theoretical}.


\subsection{Relative Likelihood}
\label{sec:background:rel_ll_and_temp_sets:relative_likelihood}
In standard maximum likelihood estimation, the goal is to find
the single best parameter that maximizes the likelihood of the
observed data. In the context of our method, we treat the decoding 
temperature $\tau \in \mathcal{T}$ as the parameter of interest.
Let $L(\tau)$ denote the likelihood of the target sequence generated 
at temperature $\tau$. Furthermore, let $\tau^{ML}$ be the optimal
temperature that yields the highest likelihood (or equivalently,
minimizes the NLL).
To characterize the plausibility of any other temperature $\tau$, we
compute its relative likelihood $\gamma(\tau)$, defined as the 
ratio of its likelihood to the maximum likelihood
\parencite{largesampledistribution, Birnbaum01061962, https://doi.org/10.2307/3315449, https://doi.org/10.1111/1467-9868.00205}:

\begin{equation}
    \gamma(\tau) = 
    \frac{L(\tau)}{L(\tau^{ML})} = 
    \frac{L(\tau)}{\sup_{\tau' \in \mathcal{T}} L(\tau')}
    \label{eq:relative_likelihood}
\end{equation}

By definition, $\gamma(\tau)$ is bounded in the interval $[0,1]$, 
where $\gamma(\tau^{ML}) = 1$. The relative likelihood provides
an intuitive measure of how well a specific temperature explains
the data compared to the optimal temperature setting.

\subsection{Constructing $\alpha$-Cuts for Temperatures}
\label{sec:background:rel_ll_and_temp_sets:alpha_cuts}
On the basis of the relative likelihood, a set of valid temperatures
can be constructed by including only those that are plausible in the  
sense of surpassing a specific threshold $\alpha \in [0,1]$. This 
set is formally referred to as an \textbf{$\alpha$-cut}
\parencite{epub31329}:

\begin{equation}
    \mathcal{T}_\alpha = 
    \{\tau \in \mathcal{T} : \gamma(\tau) \geq \alpha \}
    \label{eq:alpha_cut}
\end{equation}

According to this definition, a temperature $\tau$ is considered
implausible, and is therefore excluded from the valid set, if 
its likelihood is too small compared to the likelihood of the 
best temperature, specifically if it is less than $\alpha$
times the optimal likelihood.
